Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-09-24 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_8749e}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\caption{}\label{T_0105a}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

\section{Results: Prediction of Risk for Occult
Disease}\label{results-prediction-of-risk-for-occult-disease}

@@

\begin{itemize}
\tightlist
\item
  LNL II when the tumor extends over the mid-sagittal line
\item
  LNL III when LNL II is clinically involved, may need confirmation via
  FNA
\item
  LNL IV only when upstream level III pathologically confirmed
\item
  LNLs I and VII only when clinically involved
\item
  LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs
  metastatic, contra II, III, and IV clinically involved plus cIV via
  FNA confirmed)
\end{itemize}

@@

\caption{}\label{T_a035c}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
10 \\
early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
61 \\
advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
79 \\
advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\

>>>>

\caption{}\label{T_5ff4d}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\caption{}\label{T_f5301}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

\section{Results: Prediction of Risk for Occult
Disease}\label{sec-results-risks}

@@

\begin{itemize}
\tightlist
\item
  Irradiate LNL II when the tumor extends over the mid-sagittal line
\item
  Irradiate LNL III when LNL II is clinically involved
\item
  Irradiate LNL IV only when the upstream level III is pathologically
  confirmed via FNA
\item
  Exclude LNLs I and VII for most patients. I.e., only treat these LNLs
  when they appear clinically metastatic
\item
  Exclude LNL V in all but the most extreme cases. I.e., for advanced
  T-category tumors with midline extension, where all ipsilateral LNLs
  and the contralateral LNLs II to IV present with clinical involvement
  and the upstream involvement is additionally verified by FNA.
\end{itemize}

In essence, this is a summary of the results in
section~\ref{sec-results-risks}. However, it should be considered in
light of the limitations outlined in
section~\ref{sec-discussion-limitations} below.

\subsection{Limitations and Future
Work}\label{sec-discussion-limitations}

\subsubsection{T-Category Dependence}\label{t-category-dependence}

For some patterns of involvement the difference in the observed
prevalence for early vs advanced T-category is greater, for other
smaller, than what our model predicts. This is because only one
parameter, \(p_\text{adv.}\), governs this spread by continuing to
evolve the system a certain number of time-steps, while the speed of
spread, here the \(b_v\) parameters, remains constant. However, it is
feasible that these parameters themselves may change over time as the
tumor grows into regions with different lymph drainage pathways.

\subsubsection{Sensitivity and
Specificity}\label{sensitivity-and-specificity}

As hinted at in section~\ref{sec-data-consensus}, the literature values
for sensitivity and specificity pairs of the diagnostic modalities do
not plausibly match some of the observed data. We therefore assumed the
consensus across diagnostic modalities and pathology to represent the
ground truth. This simplification allowed us to investigate the models
general capability to describe a realistic, if not necessarily entirely
accurate dataset. But in the future, this may need to be approached more
rigorously, e.g.~by differentiating between pathologically confirmed and
clinically diagnosed involvement, as the latter is - by definition -
incapable of detecting microscopic disease.

\subsubsection{Tumor Subsites}\label{tumor-subsites}

Within the cohort of patients with oropharyngeal tumors, subgroups can
be identified whose tumors are located in different subsites, e.g.~at
the base of the tongue or on the tonsils. The lymphatic metastatic
spread patterns of these subgroups may differ slightly. If so, the model
could benefit from more fine-grained information about the precise tumor
subsite for its training and prediction. We have previously speculated
that a mixture model could be a promising approach to capture the
subsite specific spread patterns \citep{zora231470}.

@@

\caption{}\label{T_474bc}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
10 \\
early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
61 \\
advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
79 \\
advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\
